---
title: "STAT 331 Final Project"
author: "Gabe Riedel, Lena Kimura, Diego Jara, Drew Cloughley"
format: 
  html:
    code-tools: true
    toc: true
    embed-resources: true
    html-table-processing: none
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
code-fold: true
---

```{r load-packages}
#| output: false
library(tidyverse)
library(knitr)
library(gganimate)
library(gifski)
library(kableExtra)
library(glue)
library(broom)
```

# PC3 Code

```{r data-pivot}
#| output: false
income <- read_csv("data/avg_daily_income.csv")
women <- read_csv("data/mean_years_in_school_women_25_34_years.csv")

income_longer <- income |> 
  pivot_longer(cols = `1800`:`2100`, 
               names_to = "Year", 
               values_to = "Income")

women_longer <- women |> 
  pivot_longer(cols = `1970`:`2015`, 
               names_to = "Year", 
               values_to = "Number of Years in School")
```

```{r}
full_data <- income_longer |> 
  inner_join(women_longer, by = join_by(Year, country))

full_data <- full_data |> 
  mutate(Region = fct_collapse(factor(country),
    "North America" = c("Canada", "USA", "Mexico", "Belize", "Costa Rica", "El Salvador", "Guatemala", 
                        "Honduras","Nicaragua", "Panama"),
    "South America" = c("Argentina", "Bolivia", "Brazil", "Chile", "Colombia", "Ecuador", 
                        "Guyana", "Paraguay", "Peru", "Suriname", "Uruguay", "Venezuela"),
    "Caribbean" = c("Antigua and Barbuda", "Bahamas", "Barbados", "Cuba", "Dominica", 
                    "Dominican Republic", "Grenada", "Haiti", "Jamaica", "St. Lucia", 
                    "St. Vincent and the Grenadines", "Trinidad and Tobago"),
    "Europe" = c("Andorra","Austria", "Belgium", "France", "Germany", "Luxembourg", "Netherlands",
                 "Switzerland", "UK", "Belarus", "Bulgaria", "Croatia", "Czech Republic",
                 "Estonia", "Hungary", "Latvia", "Lithuania", "Moldova", "Poland", "Romania", 
                 "Russia", "Slovak Republic", "Ukraine", "Albania", "Bosnia and Herzegovina", 
                 "Cyprus", "Greece", "Italy", "Malta", "Montenegro", "Portugal", "Serbia", 
                 "Slovenia", "Spain", "North Macedonia", "Denmark", "Finland", "Iceland", 
                 "Ireland", "Norway", "Sweden"),
    "Middle East" = c("Bahrain", "Egypt", "Iran", "Iraq", "Israel", "Palestine", "Jordan", "Kuwait",
                      "Lebanon", "Oman", "Qatar", "Saudi Arabia", "UAE", "Yemen", "Turkey", "Syria",
                      "Armenia", "Azerbaijan", "Georgia"),
    "Asia" = c("Afghanistan", "Bangladesh", "Bhutan", "India", "Maldives", "Nepal", "Pakistan",
                     "Sri Lanka", "Uzbekistan", "Turkmenistan", "Tajikistan", "Kyrgyz Republic",
                     "Kazakhstan", "Brunei", "Cambodia", "Indonesia", "Lao", "Malaysia", "Myanmar",     
                     "Philippines", "Singapore", "Thailand", "Timor-Leste", "Vietnam", "China", "Japan",  
                     "North Korea", "South Korea", "Mongolia", "Taiwan"),
    "Oceania" = c("Australia", "Fiji", "Kiribati", "Marshall Islands", "New Zealand", "Papua New Guinea",
                  "Solomon Islands", "Tonga", "Vanuatu", "Micronesia, Fed. Sts.", "Samoa"),
    "Africa" = c("Algeria", "Benin", "Burkina Faso", "Cape Verde", "Chad", "Comoros", 
                 "Cote d'Ivoire", "Djibouti", "Eritrea", "Gambia", "Ghana", "Liberia", "Libya",
                 "Mali", "Mauritania", "Mauritius", "Morocco", "Niger", "Nigeria",
                 "Burundi", "Cameroon", "Central African Republic", "Congo, Dem. Rep.", 
                 "Congo, Rep.", "Ethiopia", "Equatorial Guinea", "Gabon", "Guinea",
                 "Guinea-Bissau", "Kenya", "Rwanda", "Seychelles", "Somalia", "South Sudan",
                 "Namibia", "Uganda", "Sao Tome and Principe", "Senegal", "Sierra Leone",
                 "Sudan", "Togo", "Tunisia", "Angola", "Botswana", "Eswatini", "Lesotho", "Madagascar",
                 "Malawi", "Mozambique", "South Africa", "Tanzania", "Zambia", "Zimbabwe")
    
  ))
```

# PC3 Write-up

### Data/Variable Description

In this analysis, we will use two variables, average daily income and average years in school of women ages 25-34. Our first dataset includes information on the average daily income in 2017 dollars from the year 1800 to 2100 (projected). Our second dataset includes the mean number of years enrolled in school for women between the ages 25-34 from 1970 to 2015. Each observation is uniquely identified by a different country. Our analysis will include the years shared by both datasets (1970-2015) and the shared countries (188 in common). Our datasets did not have any missing values or inconsistent data types.

### Hypothesized Variable Relationship

We hypothesize that nations with women that have spent more years in school between ages 25 and 34 will have a higher average daily income because women in this age range may have also completed graduate-level education.

# PC4

### 2.1 Data Visualization

```{r}
full_data |> 
  group_by(country) |> 
  mutate(avg_income = mean(Income), 
         avg_school = mean(`Number of Years in School`)) |>
  ggplot(aes(x = avg_school, y = avg_income, color = Region)) +
  geom_point() +
  labs(x = "Number of Years in School", 
       y = "Income", 
       title = "Income and Number of Years in School by Women across the World",
       subtitle = "Income and School Years were Averaged across Country") +
  theme_bw()
```

In the scatterplot above, every dot on the graph represents a different country where they are colored by their corresponding region. The x-axis represents the average number of years that women are in school, where the y-axis displays the average daily income for a household. To properly display all the data from the years 1970 to 2015, the average of both variables was taken respective to the county and placed on the scatterplot above.

```{r}
animation <- full_data |> 
  mutate(Year = as.numeric(Year)) |> 
  ggplot(aes(x = `Number of Years in School`, y = Income, color = Region)) +
  geom_point() +
  labs(x = "Number of Years in School", 
       y = "Income", 
       title = "Income and Number of Years in School by Women across the World 1970-2015",
       subtitle = "Income and School Years across Country and Region") +
  theme_bw()

anim <- animation + transition_time(Year) +
  ease_aes("linear")

anim

```

### 2.2 Linear Regression

$$Log_{10}(MeanIncome) = 0.3088 + 0.0959*MeanEducation $$

Because we took the base-10 logarithm of the mean income by country, a slope of 0.0959 indicates that income increases by estimate of 24.71% for every additional year of education for women ages 25-24. At 0 years of education, average income is approximately \$2.04.

```{r}
regression_model_data <- 
  full_data |> 
  group_by(country) |> 
  mutate(mean_inc = mean(Income)) |> 
  mutate(mean_educ = mean(`Number of Years in School`))

lm_income_educ <- lm(log10(mean_inc) ~ mean_educ, data = regression_model_data)
kable(tidy(lm_income_educ))
```


The $Log_{10}$ transformation of average daily income helped to linearize our model shown in the scatterplot below. Like the figure in 2.1, each dot on the graph represents a different country. The x-axis represents the average number of years that women are in school, where the y-axis displays the $Log_{10}$ of the average daily income for a household. To properly display all the data from the years 1970 to 2015, the average of both variables was taken respective to the county and placed on the scatterplot above.


```{r}
regression_model_data |> 
  ggplot(aes(x = mean_educ, 
             y = log10(mean_inc))) +
           geom_point() + 
  geom_smooth(method = lm) +
  labs(x = "Number of Years in School", 
       y = "Log10(Income)",
              title = "Log Income and Number of Years in School for Women across the World",
       subtitle = "Income and number of years in school were averaged across country") +
  theme_bw()
```

### 2.3 Model Fit

```{R}
var_response <- var(log10(regression_model_data$mean_inc))
var_fitted <- var(lm_income_educ$fitted)
var_residuals <- var(lm_income_educ$residuals)

var_table <- data.frame(c("Variance in Response", "Variance in Fitted Values", "Variance in Residuals"),
                        c(var_response, var_fitted, var_residuals))

var_table |>
  kable(col.names = c("Metric", "Variance")) |>
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover")) |>
  add_header_above(c("Response Variability" = 2)) |>
  row_spec(0, bold = TRUE, color = "black")
```

```{r}
var_prop_explained <- round(var_fitted/var_response,4) * 100
var_prop_unexplained <- round(var_residuals/var_response,4) * 100
var_percent_explained <- glue("{var_prop_explained}%")
var_percent_unexplained <- glue("{var_prop_unexplained}%")
```

The percent of variability explained by the model is `r var_percent_explained`. This means our model is accounting for a moderate amount of variability. There is still `r var_percent_unexplained` unexplained by the model, so other factors outside of years in school are influencing this response. Overall, for simple linear regression, our model is of moderate to high quality.
