---
title: "final_project"
author: "Gabe Riedel, Lena Kimura, Diego Jara, Drew Cloughley"
format: html
editor: visual
---

```{r load-packages}
library(tidyverse)
library(knitr)
```

```{r data-pivot}
income <- read_csv("data/avg_daily_income.csv")
women <- read_csv("data/mean_years_in_school_women_25_34_years.csv")

income_longer <- income |> 
  pivot_longer(cols = `1800`:`2100`, 
               names_to = "Year", 
               values_to = "Income")

women_longer <- women |> 
  pivot_longer(cols = `1970`:`2015`, 
               names_to = "Year", 
               values_to = "Number of Years in School")
```


```{r}
#| echo: false
```

#PC3 Write-up

### Data/Variable Description
  In this analysis, we will use two variables, average daily income and average years in school of women ages 25-34. Our first dataset includes information on the average daily income in 2017 dollars  from the year 1800 to 2100 (projected). Our second dataset includes the mean number of years enrolled in school for women between the ages 25-34 from 1970 to 2015. 
  Each observation is uniquely identified by a different country. Our analysis will include the years shared by both datasets (1970-2015) and the shared countries (188 in common). 

